\documentclass[a4]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{../../manual/manual}
\include{mylistings}
\makeindex

\lstset{basicstyle=\sffamily}
\author{Louise A. Dennis}

\title{Gwendolen Tutorial 3 -- More Rubble: \\
Plan Guards and Reasoning Rules}

\begin{document}
\maketitle
This is the third in a series of tutorials on the use of the \gwendolen\ programming language.  This tutorial covers the use of \prolog\ style rules as they appear in \gwendolen\ and also looks at plan guards in a little more detail.

Files for this tutorial can be found in the \texttt{mcapl} distribution in the directory \texttt{src/examples/gwendolen/tutorials/tutorial3}.

\section{Pick Up Rubble (Again)}

You will find a \gwendolen\ program in the tutorial directory called \texttt{pickuprubble\_achieve.gwen}.  Its contents should look like Listing~\ref{code:pickuprubble_achieve}.
\begin{lstlisting}[float,caption=Pick Up Rubble (Reasoning Rules),basicstyle=\sffamily,style=easslisting,language=Gwendolen,label=code:pickuprubble_achieve]
GWENDOLEN

:name: robot

:Initial Beliefs:

possible_rubble(1, 1) possible_rubble(3, 3) possible_rubble(5, 5)

:Reasoning Rules:

square_to_check(X, Y) :- possible_rubble(X, Y), ~no_rubble(X, Y);

:Initial Goals:

holding(rubble) [achieve]

:Plans:

+!holding(rubble) [achieve] : {B square_to_check(X, Y)} <- 
    move_to(X, Y);
+at(X, Y) : {~B rubble(X, Y)} <- +no_rubble(X, Y);
+rubble(X, Y): {B at(X, Y)} <- lift_rubble;
+holding(rubble): {True} <- print(done);
\end{lstlisting}

This is very similar to the second program in tutorial 2.  However instead of having 
\begin{verbatim}
{B possible_rubble(X, Y), ~B no_rubble(X, Y)}
\end{verbatim}as the guard to the first plan we have \lstinline{B square_to_check(X, Y)} as the guard.

We can then reason about whether there is a square to check using the \prolog\ style rule on line 13.  The syntax is very similar to \prolog\ syntax but there are a few differences.  We use the symbol, $\sim$, to indicate ``not'' and we can not use ``cut''s to control backtracking~\footnote{NB.  There is no reason in principle why we could no use cut, it just hasn't been implemented}.

\section{Using \prolog\ Lists}
You can use \prolog\ style list structures in \gwendolen programs.  Listing~\ref{code:pickuprubble_list} shows the previous example using lists.
\begin{lstlisting}[float,caption=Pick Up Rubble (with \prolog\ lists in Reasoning Rules),basicstyle=\sffamily,style=easslisting,language=Gwendolen,label=code:pickuprubble_list]
:name: robot

:Initial Beliefs:

possible_rubble([sq(1, 1), sq(3, 3), sq(5, 5)])

:Reasoning Rules:

square_to_check(X, Y) :- possible_rubble(L), check_rubble(L, X, Y);

check_rubble([sq(X, Y) | T], X, Y) :- ~no_rubble(X, Y);
check_rubble([sq(X, Y) | T], X1, Y1) :- no_rubble(X, Y), 
                       check_rubble(T, X1, Y1);

:Initial Goals:

holding(rubble) [achieve]

:Plans:

+!holding(rubble) [achieve] : {B square_to_check(X, Y)} <-
     move_to(X, Y);
+at(X, Y) : {~B rubble(X, Y)} <- +no_rubble(X, Y);
+rubble(X, Y): {B at(X, Y)} <- lift_rubble;
+holding(rubble): {True} <- print(done);
\end{lstlisting}

\prolog\ list structures can also be used in \gwendolen\ plans and a recursive style plan may sometimes provide a more efficient programme than the kind of program that relies on the failure of a plan to achieve a goal, to re-trigger the plan.  Listing~\ref{code:pickuprubble_list2} shows an example of this style of programming where the achieve goal calls a perform goal that recurses through the list of squares one at a time.

\begin{lstlisting}[float,caption=Pick Up Rubble (with \prolog\ lists in Plans),basicstyle=\sffamily,style=easslisting,language=Gwendolen,label=code:pickuprubble_list2]
:name: robot

:Initial Beliefs:

possible_rubble([sq(1, 1), sq(3, 3), sq(5, 5)])

:Reasoning Rules:

rubble_in_current :- at(X, Y), rubble(X, Y);

:Initial Goals:

holding(rubble) [achieve]

:Plans:

+!holding(rubble) [achieve] : {B possible_rubble(L)} <- 
    +! check_all_squares(L) [perform];

+!check_all_squares([]) [perform] : {True} <- print(done);
+!check_all_squares([sq(X, Y) | T]) : {~B rubble_in_current} <- 
    move_to(X, Y), 
    +!check_all_squares(T) [perform];
+!check_all_squares([sq(X, Y) | T]) : {B rubble_in_current} <- 
    print(done);

+at(X, Y) : {~B rubble(X, Y)} <- +no_rubble(X, Y);

+rubble(X, Y): {B at(X, Y)} <- lift_rubble;
\end{lstlisting}

\section{More Complex \prolog\ Reasoning -- Grouping predicates under a negation}

\texttt{pickuprubble\_grouping.gwen} shows more complex use of Reasoning Rules including some syntax not available in \prolog.  This is shown in listing~\ref{code:pickuprubble_grouping}
\begin{lstlisting}[float,caption=Pick Up Rubble (More Complex Reasoning Rules),basicstyle=\sffamily,style=easslisting,language=Gwendolen,label=code:pickuprubble_grouping]
GWENDOLEN

:name: robot

:Initial Beliefs:

possible_rubble(1, 1)
possible_rubble(3, 3)
possible_rubble(5, 5)

:Reasoning Rules:

square_to_check(X, Y) :- possible_rubble(X, Y), ~no_rubble(X, Y);
done :- holding(rubble);
done :- ~ (possible_rubble(X, Y), ~no_rubble(X, Y));

:Initial Goals:

done [achieve]

:Plans:

+!done [achieve] : {B square_to_check(X, Y)} <- move_to(X, Y);

+at(X, Y) : {~B rubble(X, Y)} <- +no_rubble(X, Y);

+rubble(X, Y): {B at(X, Y)} <- lift_rubble;

+holding(rubble): {True} <- print(done);
\end{lstlisting}

In this program the agent's goal is to achieve \lstinline{done}.  It achieves this either if it is holding rubble (deduced using the code on line 14), or if there is no square it thinks may possibly contain rubble that has no rubble in it (deduced using the code on line 15).

\begin{sloppypar}
The rule on line 15
\begin{verbatim}
done :- ~ (possible_rubble(X, Y), ~no_rubble(X, Y));
\end{verbatim}
isn't standard \prolog\ syntax.  Here we group the two predicates \lstinline{possible_rubble(X, Y), ~no_rubble(X, Y)} (from \lstinline{square_to_check}) together using brackets and then negate the whole concept (i.e., there are no squares left to check).
\end{sloppypar}

\subsection{Some Simple Exercises to Try}
\begin{enumerate}
\item Try removing the initial belief that there is possible rubble in square (5, 5).  You should find the the program still completes and prints out done.
\item Try replacing the rule on line 15 with one that refers to \lstinline{square_to_check}
\end{enumerate}



\section{Using Goals in Plan Guards}

\texttt{pickuprubble\_goal.gwen} shows how reasoning rules can be used to reason about both goals and beliefs.  This is shown in listing~\ref{code:pickuprubble_goal}
\begin{lstlisting}[float,caption=Pick Up Rubble (Goals in Guards),basicstyle=\sffamily,style=easslisting,language=Gwendolen,label=code:pickuprubble_goal]
GWENDOLEN

:name: robot

:Initial Beliefs:

possible_rubble(1, 1)
possible_rubble(3, 3)
possible_rubble(5, 5)

:Initial Goals:

rubble(2, 2) [achieve]

:Plans:

+!rubble(2, 2) [achieve]: {True} <- +! holding(rubble)[achieve],
    move_to(2, 2), 
    drop;

+!holding(rubble) [achieve] : 
    {B possible_rubble(X, Y), ~B no_rubble(X, Y)} <- move_to(X, Y);

+at(X, Y) : {~B rubble(X, Y)} <- +no_rubble(X, Y);

+rubble(X, Y): {B at(X, Y), G holding(rubble) [achieve]} <- 
    lift_rubble;
\end{lstlisting}
Recall that in the exercises at the end of tutorial 2 we had to use a belief to prevent the robot picking up the rubble after it had put it down.  Here instead we have added \lstinline{G holding(rubble) [achieve]} as a guard to the plan that is activated when the robot sees some rubble.  In this case it only picks up the rubble if it has goal to be holding rubble.

\section{Reasoning about Beliefs and Goals}
\texttt{pickuprubble\_goalat.gwen} shows how goals can be used in plan guards.  This is shown in listing~\ref{code:pickuprubble_goalat}
\begin{lstlisting}[float,caption=Pick Up Rubble (Reasoning about Goals),basicstyle=\sffamily,style=easslisting,language=Gwendolen,label=code:pickuprubble_goalat]
GWENDOLEN

:name: robot

:Initial Beliefs:

possible_rubble(1, 1)
possible_rubble(3, 3)
possible_rubble(5, 5)

:Reasoning Rules:

rubble_at_22 :- holding(rubble), at(2, 2);

:Initial Goals:

at(2, 2) [achieve]

:Plans:

+!at(X, Y) [achieve]: {True} <- +! holding(rubble)[achieve],
     move_to(X, Y);

+!holding(rubble) [achieve] : 
    {B possible_rubble(X, Y), ~B no_rubble(X, Y)} <- move_to(X, Y);

+at(X, Y) : {~B rubble(X, Y), ~B rubble_at_22} <- 
    +no_rubble(X, Y);
+at(X, Y) : {B rubble_at_22} <- drop;

+rubble(X, Y): {B at(X, Y), G rubble_at_22 [achieve]} <- 
    lift_rubble;
\end{lstlisting}
Here the reasoning rule on line 13 is used both in the plan on line 29 in order to reason about whether the robot is holding the rubble at square (2, 2) \emph{and} in the plan on line 31 to deduce that the robot has a goal to get the rubble to square (2, 2), from the the fact that it has a goal to be holding rubble (added on line 21) \emph{and} a goal to be at (2, 2) -- the initial goal.

Note that we can't use reasoning rules to break down a goal into subgoals.  So if you gave the robot the initial goal \lstinline{rubble_at_22} you need to provide a plan specifically for \lstinline{rubble_at_22}.  It is no good providing a plan for holding rubble and a plan for being at 2, 2 and then expecting the robot to compose these sensibly in order to achieve \lstinline{rubble_at_22}.

Try change the agents initial goal to \lstinline{rubble_at_22 [achieve]} without changing anything else in the program.  You should see a warning generated that the agent can not find a plan for the goal.  At this point the program will fail to terminate (when \gwendolen\ can't find a plan for a goal it cycles infinitely looking a plan to handle a failed goal (most programs don't include one of these).  {\bf You will need to terminate the program} (control-C at the command line or clicking the red stop square in Eclipse).

\section{Some Simple Programs to Write}
NB.  The environment \texttt{gwendolen.tutorials.SearchAndRescueEnv} contains rubble both at (5, 5) and at (3, 4).
\begin{enumerate}
\item Write a program to make the robot check every square in a 5x5 grid (i.e., (1, 1), (1, 2), (1, 3) etc.,) until it finds some rubble at which point it stops.  Try implementing this program both with and without using lists in plans.  (NB.  For the list version you may need to insert a plan that asserts a belief when the rubble is seen, in order to make sure the robot doesn't progress through the squares to rapidly.  See comment about \lstinline{do_nothing} in the next exercise and further discussion of this in later tutorials.
\item Write a program to make the robot search every square in a 5x5 grid (i.e., (1, 1), (1, 2) etc.,) taking all the rubble it finds to the square (2, 2) until it believes there is only rubble in square (2, 2).

Hints:
\begin{enumerate}
\item You may see the warning similar to:
\begin{small}
\begin{verbatim}
ail.semantics.operationalrules.GenerateApplicablePlansEmptyProblemGoal[WARNING|main|2:09:29]: 
Warning no applicable plan for goal _aall_squares_checked() 
\end{verbatim}
\end{small}
As noted above, this warning appears if the agent can not find a plan to achieve a goal.  Sometimes this arises because of bugs in the code, but it can also happen if the agent has not had a chance to process all new perceptions/beliefs before it once again looks for a plan to achieve a goal (we will talk about this some more in later tutorials).  

It may be worth adding an action \lstinline{do_nothing} into your plan, this will act to delay the next time the agent attempts to achieve the goal giving it time to process all new beliefs.
\item You may need to include \lstinline{at(2, 2)} in your goal in some way to make sure the agent actually takes the final piece of rubble to the square (2, 2).
\end{enumerate}
\end{enumerate}
\begin{sloppypar}
Sample answers for these two exercises can be found in \texttt{gwendolen/examples/tutorials/tutorial3/answers}.
\end{sloppypar}

\end{document}
