\label{tutorial:gwendolen:bda}
This is the second in a series of tutorials on the use of
the \gwendolen\ programming language\index{Gwendolen}.  This tutorial
covers the basics of beliefs\index{belief}, goals\index{goal} and
actions\index{action} as they appear in \gwendolen.

Files for this tutorial can be found in the \texttt{mcapl}
distribution in the directory
\begin{quote}
\texttt{src/examples/gwendolen/tutorials/tutorial2}.
\end{quote}

\section{Pick Up Rubble}

\begin{sloppypar}
You will find a \gwendolen\
program\index{Gwendolen}\index{example!pickuprubble} in the tutorial
directory called \texttt{pickuprubble.gwen}.  It's contents should
look like Example~\ref{code:pickuprubble}.
\end{sloppypar}
\begin{ourexample}
\label{code:pickuprubble}\quad \\
\begin{lstlisting}[basicstyle=\sffamily,language=Gwendolen,style=easslisting]
GWENDOLEN

:name: robot

:Initial Beliefs:

:Initial Goals:

goto55 [perform]

:Plans:

+!goto55 [perform] : {True} <- move_to(5, 5);

+rubble(5, 5): {True} <- lift_rubble;

+holding(rubble): {True} <- print(done);
\end{lstlisting}
\end{ourexample}
This is a program for moving around a simple grid based
environment\index{environment} and picking up rubble.  The robot can
perform three actions\index{action} in this
environment, \index{Gwendolen}\index{example!pickuprubble}
\begin{description}
\item[move\_to(X, Y)] moves to grid square (X, Y) and adds the belief\index{belief}\index{belief!add} \lstinline{at(X, Y)}.
\begin{sloppypar}
\item[lift\_rubble] attempts picks up a piece of rubble and adds the belief\index{belief}\index{belief!add} \lstinline{holding(rubble)} if there is rubble at the robot's location.
\end{sloppypar}
\item[drop] drops whatever the robot is holding and removes any beliefs\index{belief}\index{belief!remove} about what the robot is holding.
\end{description}
The default actions\index{action}\index{action!print}
(e.g., \lstinline{print}) are also available to the robot.  As the
environment\index{environment} is set up there is a block of rubble at
square (5, 5) which the robot will see if is in square (5, 5).  When
the robot picks something up it can see that it is holding it.  This
environment is programmed in \java\ and is the
class \lstinline{gwendolen.tutorials.SearchAndRescueEnv}\index{SearchAndRescueEnv
(class)}.

The program can be understood as follows.  \index{Gwendolen}
\begin{description}
\item[Line 1] states the language in which the program is written (this is because the \ail\ allows us to create multi-agent systems from programs written in several different languages).  \index{Gwendolen}
\item[Line 3] gives the name of the agent (\lstinline{robot}).  \index{Gwendolen!agent name}
\item[Line 5] starts the section for initial beliefs (there are none).  \index{belief}\index{belief!initial}
\item[Line 7] starts the section for initial goals.  There is one a \emph{perform} goal to \lstinline{goto55}.\index{goal}\index{goal!initial}
\item[Line 11] starts the section for plans\index{plan}.  There are
  three plans.  The first (line 13) states that in order to
  perform\index{goal}\index{goal!perform} \lstinline{goto55} the agent
  must move to square (5, 5).  The second (line 15) states that if the
  agent sees rubble at 5, 5 it should lift the rubble and the third
  (line 17) states that if the agent sees it is holding rubble then it
  should print done.\index{action}\index{action!print}
\end{description}

There are three different sorts of syntax being used here to
distinguish between beliefs, goals and
actions.\index{Gwendolen}\index{belief}\index{goal}\index{action}

\begin{description}
\item[Beliefs]\index{belief} Beliefs are predicates (e.g.,
  \lstinline{rubble(5, 5)} that are preceded either by a + symbol (to
  indicate a adding a belief\index{belief!add}) or a - symbol (to
  indicate removing a belief\index{belief!remove}).
\item[Goals]\index{goal} Goals are predicates preceded by an
  exclamation mark (e.g., \lstinline{!goto55}) again these are then
  preceded either by a + symbol (to indicate a adding a
  goal\index{goal!add}) or a - symbol (to indicate removing a
  goal\index{goal!remove}).  After the goal predicate there is also a
  label stating what kind of goal it is.  Goals can either be
  \emph{perform}\index{goal!perform} goals or
  \emph{achieve}\index{goal!achieve} goals.  We will discuss the
  difference between these in a moment.
\item[Actions]\index{action} Are just predicates.  Actions are
  performed externally to the agent and can not be added or removed
  (they are just done).
\end{description}

\subsection{Running the Program and Getting more Log output}\index{Gwendolen}\index{Gwendolen!running programs}

\begin{sloppypar}
To run the program you need to call \texttt{ail.mas.AIL}\index{AIL
  (class)} and supply it with a suitable configuration
file\index{configuration!gwendolen}\index{Gwendolen!configuration}.
You will find an appropriate configuration file,
\texttt{pickuprubble.ail}\index{example!pickuprubble} in the same
directory as \texttt{pickuprubble.gwen}.  You can do this either from
the command line or using the IntelliJ or Eclipse \texttt{run-AIL}\index{run-AIL}
configuration as detailed in the \mcapl\ manual.
\end{sloppypar}

Run the program now.\index{Gwendolen}\index{Gwendolen!running programs}

As in Tutorial 1, all you see is the robot
printing\index{action}\index{action!print} the message
\lstinline{done} once it has finished.  However what has happened is
that the robot moved to square (5, 5) (because of the perform
goal\index{goal}\index{goal!perform}).  Once in square (5, 5) it saw
the rubble and so lifted it (thanks to the second plan\index{plan}).
Once it had lifted the rubble it saw that it was holding it and
printed \lstinline{done} (thanks to the third plan).

You can get more information about the execution of the program by
changing the logging information in the configuration file.  Open the
configuration file and
edit\index{Gwendolen}\index{logging}\index{Gwendolen!logging}\index{configuration!gwendolen}\index{Gwendolen!configuration}
\begin{verbatim}
log.warning = ail.mas.DefaultEnvironment
\end{verbatim}
to
\begin{verbatim}
log.info = ail.mas.DefaultEnvironment
\end{verbatim}
You will now see logging information printed out about each action the robots takes.\index{DefaultEnvironment (class)}

If you add the line
\begin{verbatim}
log.format = brief
\end{verbatim}
to the configuration file you will get the log messages in a briefer form.\index{log.format}\index{Gwendolen}\index{Gwendolen!configuration}\index{configuration!gwendolen}

\section{Perform and Achieve Goals}\index{Gwendolen}\index{goal}\index{goal!perform}\index{goal!achieve}
\lstinline{pickuprubble_achieve.gwen} is a slightly more complex
version of the rubble lifting program which introduces some new
concepts.  It is shown in
Example~\ref{code:pickuprubble_achieve}\index{example!pickuprubble}
\begin{figure}[htb]
\begin{ourexample}
\label{code:pickuprubble_achieve} \quad \\
\begin{lstlisting}[basicstyle=\sffamily,style=easslisting,language=Gwendolen]
GWENDOLEN

:name: robot

:Initial Beliefs:

possible_rubble(1, 1)
possible_rubble(3, 3)
possible_rubble(5, 5)

:Initial Goals:

holding(rubble) [achieve]

:Plans:

+!holding(rubble) [achieve] : 
  {B possible_rubble(X, Y), ~B no_rubble(X, Y)} <- move_to(X, Y);

+at(X, Y) : {~B rubble(X, Y)} <- +no_rubble(X, Y);

+rubble(X, Y): {B at(X, Y)} <- lift_rubble;

+holding(rubble): {True} <- print(done);
\end{lstlisting}
\end{ourexample}
\end{figure}
The first changes are in lines 7-9.  Here we have a list of initial
beliefs\index{belief}\index{belief!initial}.  The agent believes there
may be rubble in one of three squares (1, 1), (3, 3) and (5, 5).  As
we know, in the environment, there is only rubble in (5,
5).\index{environment}

The next change is in line 13.  Here instead of a \emph{perform}
goal\index{goal}\index{goal!perform}, \lstinline{goto55 [perform]}
there is an \emph{achieve} goal\index{goal!achieve},
\lstinline{holding(rubble) [achieve]}.  The difference between perform
goals and achieve goals is as follows:
\begin{itemize}
\item When an agent adds a perform goal it searches for a plan for
  that goal, executes the plan and then drops the goal, it does not
  check that the plan has succeeded.\index{goal!perform}
\item When an agent adds an achieve goal it searches for a plan for
  that goal, executes the plan and then checks to see if it now has a
  belief corresponding to the goal.  If it has no such beliefs it
  searches for a plan again, if it does have such a belief then it
  drops the goal.\index{goal!achieve}
\end{itemize}
In the case of this program the robot will continue executing the plan
for \lstinline{holding(rubble)} until it actually believes that it is
holding some rubble.\index{belief}

On lines 17 and 18, you can see the plan\index{plan} for the
goal\index{goal}.  This plan no longer has a trivial
guard\index{plan!guard}.  Instead the plan only applies if the agent
believes\index{belief} that there is possible rubble in some square
(X, Y) and it does not (the $\sim$ symbol) believe\index{belief!not
  believe} there is no rubble in that square.  If it can find such a
square then the robot moves\index{action} to it.  The idea is that the
robot will check each of the possible rubble squares in turn until it
successfully finds and lifts some rubble.  Note that we are using
capital letters for variables\index{Gwendolen!variables} that can be
unified against beliefs (like in \prolog).

The plan\index{plan} at line 20 gets the robot to add the
belief\index{belief} \lstinline{no_rubble(X, Y)} if it is at some
square, (X, Y), and it can't see any rubble there.  By this means the
plan in lines 17 and 18 will be forced to pick a different square next
time it executes.  Up until now all the plans you have used have
simply executed actions in the plan body.  This one adds a
belief.\index{Gwendolen}

The plan at line 22 is similar to the plan in line 15 of
Example~\ref{code:pickuprubble} only in this case we are using
variables for the rubble coordinates rather than giving it the
coordinates (5, 5).\index{Gwendolen}\index{example!pickuprubble}

You can run this program using the \lstinline{pickuprubble_achieve.ail} configuration file.\index{Gwendolen}

\section{Some Simple Exercise to Try}\index{Gwendolen!exercises}
\begin{enumerate}
\item Instead of having the robot print \lstinline{done} once it has
  the rubble, get it to move to square (2, 2) and drop the rubble.

Hint: you may find you need to add a ``housekeeping'' belief that the
rubble has been moved to prevent the robot immediately picking up the
rubble once it has been dropped.
\item Rewrite the program so that instead of starting with an
  achievement goal \lstinline{holding(rubble)}, it starts with an
  achievement goal \lstinline{rubble(2, 2)} -- i.e., it wants to
  believe there is rubble in square (2, 2).

Hint: you may want to reuse the plan for achieve
\lstinline{holding(rubble)} by setting it up as a
subgoal\index{goal}\index{goal!subgoal}.  You can use this by adding
the command \lstinline{+!holding(rubble) [achieve]} in the body of a
plan.
\end{enumerate}
\begin{sloppypar}
Sample answers for these two exercises can be found in \texttt{gwendolen/examples/tutorials/tutorial2/answers}.
\end{sloppypar}


